### 2020-05-01 Writing/disparity_review.tex
added color tags for conversion

\textcolor{blue}{We have removed hyphens from ``disparity-through-time''}

@@

\textcolor{blue}{We changed the sentence to:}

@@

\textcolor{blue}{We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:}

@@

\textcolor{blue}{And added a brief mention of protocols for axis selection:}

@@

% NC: Not sure if this really answers the issue they had. Do we mean that categorical are always more problematic???

\textcolor{blue}{We've now added a reference to the Cailliez (1983) correction to remedy for euclideanarity}:

\textit{although this can sometimes be corrected, Cailliez 1983} l.@@@

@@

\textcolor{blue}{We have added mentions of other tests common in morphological disparity analyses:}

@@

\textcolor{blue}{We suspect this is probably due to us doing a poor job at explaining how these sections differ. In a standard disparity workflow, we tend to first extract disparity values for our chosen subsets of taxa which we refer to as "Measuring disparity". Following from this, some (though not all) studies go on to determine whether differences among subsets are statistically different. We refer to this step as "Testing hypotheses". To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity. We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.}

\textcolor{blue}{For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).}
% TG: Not sure if that's enough but lacking inspiration for now
% NC: I think the error the reviewer made is what I've written above, so changing the names should deal with it.

@@

\textcolor{blue}{We've now added the following caveats to this section:}

@@

\textcolor{blue}{We appreciate the reviewer's suggestion here and tried hard to implement it. Unfortunately it led to the sections being even more confused as many of our points apply across the four types of disparity analysis. We also didn't feel we had stuck to this structure in the earlier sections, so the structure here didn't make as much sense as the reviewer assumed it might. We therefore reverted back to the original structure.}

% TG: not sure what's meant here
% NC: What they mean is can we use the 4 types of disparity analysis from the intro to say: For this type of analysis you should do this test, for this type do this test. But I don't think we can...

@@

\textcolor{blue}{We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:}

\textit{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of \citealt{edgar1935irises,fisher1936use}).
Morphospaces: different mathematical representations of a morphospace. A trait matrix can be an ordinated matrix (e.g. in \citealt{tyler2011detecting} - or transformed into a distance matrix \citealt{Close2015}, not represented here). % NC: Not sure the brackets are in the right places here? As is the sentence reads "A trait matrix can be an ordinated matrix (....)."
Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
Visualisation: different  ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}

@@

\textcolor{blue}{This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:}

@@

\textcolor{blue}{We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.}

@@

\textcolor{blue}{We removed the  ``etc.''}

@@

\textcolor{blue}{We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:}

@@

\textcolor{blue}{We've added the term ``geometric'' to be more explicit.}

@@

\textcolor{blue}{We've changed the sentence to:}

@@

\textcolor{blue}{The following three points by the reviewer are closely linked so we deal with them together}

@@

\textcolor{blue}{In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests. We have also, as suggested by the reviewer, shortened the other paragraphs and expanded this section. This is a difficult section to write because the possible statistical tests heavily rely on the kind of data available and the question being asked. In many studies only visual observations of differences are made without statistical testing. The edited text is as follows:}

@@

\textcolor{blue}{We appreciate the reviewer's point about inattentive readers missing key sections. We hope that we have solved this in part by renaming these sections to more accurately represent their content (changed to 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity). As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).}

\textcolor{blue}{We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper. We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.}

@@

\textcolor{blue}{We agree this is a really important point that we have not expnaded on previously du to worries about the word count. We've now added a caveat to this section, as mentioned in the main comments:}

@@

\textcolor{blue}{We've also added the following to emphasize this point:}

@@

	\item \textbf{\textcolor{blue}{Disparity through time}.} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?

@@

	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles \citep{Pierce2008, Friedman2010}.
	The links between form and function, however, are not always clear.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
    \textcolor{blue}{It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct} \citep{Wainwright2005}.

@@

% \begin{figure}[!htbp]
% \centering
%    \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
% \caption{
%     The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{\textcolor{blue}{disparity through time}} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
%     These categories are not independent and many disparity studies will cover more than one.
% }
% \label{Fig:disparity}
% \end{figure}

@@

The points above assume that researchers are collecting their own data for disparity analyses, but this is often not the case.
Discrete characters are commonly recycled from phylogenetic studies (e.g. \citealt{Brusatte2008,Close2015}).
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
This needs to be considered when interpreting results, especially \textcolor{blue}{as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).}
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
The influence of trait choice on resulting disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018}.

@@

Disparity analyses often use ordination techniques for dimension reduction.
Ordinations are statistical methods that map observed variables onto a new space of reduced dimension while maintaining the requirement that similar observations are closer together than dissimilar ones (e.g. principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS).
They come in many flavours depending on the data and the desired morphospace properties.
For example, quantitative (continuous) data can be reduced using PCA, and \textcolor{blue}{dissimilarity matrices based on qualitative, quantitative or mixed data types} can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
\textcolor{blue}{Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace} \citep{lehmann2019}. 
\textcolor{blue}{Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.}

@@

Like most other aspects of disparity analyses, however, reducing dimensionality can be fraught.
In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{geometric} morphometric data \citep{Deline2018, Gerber2017}.

Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question (for example, if the first few ordination axes represent elongation of the beak in birds, but the question is about wing disparity).
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data are non-Euclidean (e.g. some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
This is problematic when comparing the position of groups in multidimensional space, as the distances might not be linear \textcolor{blue}{(although this can sometimes be corrected \citealt{cailliez1983})}.
Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
Although some gradients are easy to detect or interpret (e.g.
the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not (e.g. \citealt{Weisbecker2019}).
For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0 has no biological meaning if these are binary alternate states.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space2.pdf}
\caption{
    \textcolor{blue}{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of} \citealt{edgar1935irises,fisher1936use}).
    \textcolor{blue}{Morphospaces: different mathematical representations of a morphospace. A trait matrix can be ordinated matrix (e.g. in }\citealt{tyler2011detecting}\textcolor{blue}{ - or transformed into a distance matrix }\citealt{Close2015}\textcolor{blue}{, not represented here).}
    \textcolor{blue}{Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
    Visualisation: different  ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}
}
\label{Fig:morphospace}
\end{figure}

\textcolor{blue}{\subsection{Summarising disparity using disparity indices} \label{section:metrics}}

@@

When considering only one dimension, \textcolor{blue}{disparity indices can be used to compare the spread of distributions (e.g. the range, quantiles or variance) or the differences in the central tendencies (i.e. mean, median or mode) of groups in the morphospace.}
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers.
Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if the goal is to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g.
volume; \citealt{Diaz2016}, distance from the centroid; \citealt{Hopkins2017, Finlay2015}, variance and range; \citealt{Brusatte2008}).
Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g.
is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015} - see section \ref{section:ordination} above).

@@

\textcolor{blue}{\subsection{Testing for differences in disparity} \label{section:testing}}

No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018}, but some of these advances have yet to be implemented in disparity analyses.
Instead, hypothesis testing has been mostly confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance (PERMANOVA) \citep{Anderson2001} that tests whether two groups share the same centroid and dispersion based on a distance matrix between observations.
\textcolor{blue}{The last decade has also seen a series of developments based on this test (e.g. the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.}

It is also important to consider which data should be subjected to a statistical test.
For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
This has two advantages: (i) when the disparity index is unidimensional (e.g. the sum of variances), bootstrapping the data generates a distribution of the index that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
However, bootstrapped data are pseudoreplicates and thus non-independent \textcolor{blue}{and can increase the false positive rate (Type I error)} \citep{strube1988bootstrap}.
\textcolor{blue}{This, again, highlights the importance of tailoring the statistical test to the data and question at hand.}

@@

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998}.
Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common), the non-independence between observations should be taken into account.
It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axes of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
\textcolor{blue}{Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution. Standard methods assume a model of Brownian motion, i.e. a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution. In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}. If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.}

One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{disparity through time} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic,Guillerme2018}.
Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e. the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e. the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g. the ordination scores; \citealt{lloyd2018}).

>>>>

\textcolor{blue}{COLORTAGIN[We have removed hyphens from ``disparity-through-time'']COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We changed the sentence to:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've added that PCO can be used on dissimilarity matrices built using qualitative, or quantitative data, or mixed data types:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[And added a brief mention of protocols for axis selection:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've now stated that it is not always problematic and added a reference to the Cailliez (1983) correction to remedy for euclideanarity]COLORTAGOUT}:

\textit{This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected; Cailliez 1983} l.@@@

@@

\textcolor{blue}{COLORTAGIN[We have added mentions of other tests common in morphological disparity analyses:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We suspect this is probably due to us doing a poor job at explaining how these sections differ. In a standard disparity workflow, we tend to first extract disparity values for our chosen subsets of taxa which we refer to as "Measuring disparity". Following from this, some (though not all) studies go on to determine whether differences among subsets are statistically different. We refer to this step as "Testing hypotheses". To reduce confusion we have renamed these sections as 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity. We have retained the Disparity and phylogeny subsection as this is somewhat of a side issue that is important for these studies, but depending on the clade in question might be dealt with well or might be ignored due to lack of phylogenies for the group that could actually be used for these analyses.]COLORTAGOUT}

\textcolor{blue}{COLORTAGIN[For the Testing section, we've added some references to specific tests and insisted throughout the section that it is important to chose the right statistical test according to the data and question at hand (and pointed to some general reviews detailing some of the options - see response to reviewer 1 above and response to specific comments below).]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've now added the following caveats to this section:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We appreciate the reviewer's suggestion here and tried hard to implement it. Unfortunately it led to the sections being even more confused as many of our points apply across the four types of disparity analysis. We also didn't feel we had stuck to this structure in the earlier sections, so the structure here didn't make as much sense as the reviewer assumed it might. We therefore reverted back to the original structure.]COLORTAGOUT}

\textcolor{blue}{COLORTAGIN[If we understand this reviewer well, they suggest we partition the methodological section per type of disparity analysis (e.g. for descriptive disparity analysis, one could apply such and such tests, etc...). If this is the case, we unfortunately think that this will make the paper to prescriptive and probably more complex to read since methods can be applied to different types of analysis.]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We have now updated the figure to be just illustrate the path from raw data to PCA (illustrating that both can be morphospaces). We updated the caption accordingly:]COLORTAGOUT}

\textit{Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of \citealt{edgar1935irises,fisher1936use}).
Morphospaces: different mathematical representations of a morphospace. A trait matrix can be an ordinated matrix (e.g. in \citealt{tyler2011detecting}) or transformed into a distance matrix (e.g. in \citealt{Close2015}, not represented here).
Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
Visualisation: different  ways to represent the morphospace in 2D.
Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).}

@@

\textcolor{blue}{COLORTAGIN[This is a very fair observation. We've changed the sentence mentioning disparity and ecology and palaeobiology as follows:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We phrased the paragraph as shown below. Hopefully this has improved the flow to the main point.]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We removed the  ``etc.'']COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've added the following sentence highlighting the effect of the chosen distance metric and referring to a great paper looking at it in more detail:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've added the term ``geometric'' to be more explicit.]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've changed the sentence to:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[The following three points by the reviewer are closely linked so we deal with them together]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[In response to this, and comments by reviewer 1, we have fleshed out the "Testing" section to mention other existing tests. We have also, as suggested by the reviewer, shortened the other paragraphs and expanded this section. This is a difficult section to write because the possible statistical tests heavily rely on the kind of data available and the question being asked. In many studies only visual observations of differences are made without statistical testing. The edited text is as follows:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We appreciate the reviewer's point about inattentive readers missing key sections. We hope that we have solved this in part by renaming these sections to more accurately represent their content (changed to 1) Summarising disparity using disparity indices and 2) Testing for differences in disparity). As mentioned above, we tried reshuffling some of this material, but the current structure best suits the way most disparity analyses are planned and carried out (in our experience).]COLORTAGOUT}

\textcolor{blue}{COLORTAGIN[We avoid giving general advice because the idea with this paper was to inform but to be as non-prescriptive as possible (our take home message after the meeting at the origin of the paper was: ``do whatever disparity analysis you want but do it for the right reasons'') so we already had to tone down some prescriptive aspects in previous draft versions of this paper. We have however, highlighted the importance of applying the appropriate test to for the question and data twice in this section to really emphasize the point.]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We agree this is a really important point that we have not expnaded on previously du to worries about the word count. We've now added a caveat to this section, as mentioned in the main comments:]COLORTAGOUT}

@@

\textcolor{blue}{COLORTAGIN[We've also added the following to emphasize this point:]COLORTAGOUT}

@@

	\item \textbf{\textcolor{blue}{COLORTAGIN[Disparity through time}.]COLORTAGOUT} This approach investigates how the morphologies of organisms have changed over time, by focussing on the disparity of taxa in particular time intervals or slices.
	This approach has been used widely in palaeobiology to answer a range of macroevolutionary questions, such as: how does disparity accumulate over the history of a clade \citep{prentice2011evolution, Guillerme2018}, or how does disparity change up to and across mass extinction events\citep{Friedman2010}?

@@

	\item \textbf{Disparity as a proxy for ecology.} The disparity of a group can be used as a proxy for either the functional role it plays within an ecosystem or its ecological niche.
	This approach assumes that groups with high disparity are also likely to be functionally and ecologically diverse, and that groups found in similar regions of shape space will have similar functional and ecological roles \citep{Pierce2008, Friedman2010}.
	The links between form and function, however, are not always clear.
	Traits can be linked to multiple functions and multiple functions can be linked to a single trait \citep{Wainwright2005}.
	This approach has been used to investigate hypotheses of competitive replacement \citep{tyler2011detecting} and changes in ecosystem function during and after mass extinctions \citep{Friedman2010}.
    \textcolor{blue}{COLORTAGIN[It is one of the primary ways to investigate ecosystem functioning in palaeobiology when the study species (and their functional characteristics) are extinct]COLORTAGOUT} \citep{Wainwright2005}.

@@

% \begin{figure}[!htbp]
% \centering
%    \includegraphics[width=0.9\textwidth]{Figures/figure_disparities.pdf}
% \caption{
%     The four main types of disparity analysis. \textbf{Descriptive disparity} focuses on describing the features of morphospace occupation; \textbf{\textcolor{blue}{COLORTAGIN[disparity through time}} investigates the evolution of the morphospace through time including the effect of extinction events; \textbf{disparity and taxonomic diversity} compares different measures of biodiversity; \textbf{disparity as a proxy for ecology} uses disparity as a proxy for the ecological or functional role of a group.
%     These categories are not independent and many disparity studies will cover more than one.
% }
% \label{Fig:disparity}
% \end{figure}

@@

The points above assume that researchers are collecting their own data for disparity analyses, but this is often not the case.
Discrete characters are commonly recycled from phylogenetic studies (e.g. \citealt{Brusatte2008,Close2015}).
This is an efficient approach to character sampling, but it may artifactually increase disparity between phylogenetically distinct groups, since phylogenetic characters are often collected to discriminate among groups.
This needs to be considered when interpreting results, especially \textcolor{blue}{COLORTAGIN[as synapomorphies can lead to apparent shifts or increases in disparity when new clades appear (especially if the character-state distribution is skewed towards a particular clade).]COLORTAGOUT}
Furthermore, many datasets are limited to subsets of anatomy that are at least implicit samples of overall anatomy, but explicit tests of this assumption have shown that different aspects of morphology can exhibit different patterns of disparity \citep{Hopkins2017}.
The influence of trait choice on resulting disparity patterns can be especially challenging where the available data has non-random missing anatomical parts, such as the absence of soft tissue in the fossil record \citep{Deline2018}.

@@

Disparity analyses often use ordination techniques for dimension reduction.
Ordinations are statistical methods that map observed variables onto a new space of reduced dimension while maintaining the requirement that similar observations are closer together than dissimilar ones (e.g. principal component analysis - PCA; principal coordinates analysis - PCO/PCoA; non-metric multidimensional scaling - NMDS).
They come in many flavours depending on the data and the desired morphospace properties.
For example, quantitative (continuous) data can be reduced using PCA, and \textcolor{blue}{COLORTAGIN[dissimilarity matrices based on qualitative, quantitative or mixed data types]COLORTAGOUT} can be reduced using PCO (which is equivalent to metric multidimensional scaling (MDS)) or non-metric multidimensional scaling (NMDS; see \citealt{Legendre2012} chapter 9 for a detailed overview of ordination methods and properties).
\textcolor{blue}{COLORTAGIN[Note that for PCO, the distance metric used can have significant impacts on the resulting morphospace]COLORTAGOUT} \citep{lehmann2019}. 
\textcolor{blue}{COLORTAGIN[Choice of distance metric is therefore crucial, and should not be overlooked when using PCO.]COLORTAGOUT}

@@

Like most other aspects of disparity analyses, however, reducing dimensionality can be fraught.
In the case of ordination, subsampling axes from the ordination can lead to misinterpretation of the results.
Although a common technique is to consider the \textit{d} axes that encompass 95\% or 99\% of the variance in the dataset \textcolor{blue}{COLORTAGIN[(either by manually selecting the \textit{d} axes that encompasses the desired cumulative variance or using methods such as the broken stick model; \citealt{Legendre2012} p.410]COLORTAGOUT)}, the interpretation of these principal axes can be unrepresentative of the data and lead to misinterpretation of the biological variation mapped on these axes \citep{Bookstein2015, Weisbecker2019}.
Visual interpretations of multidimensional data can be particularly misleading, not least as multidimensional spaces are not necessarily Euclidean even when analysing \textcolor{blue}{COLORTAGIN[geometric]COLORTAGOUT} morphometric data \citep{Deline2018, Gerber2017}.

Interpreting biological variation along the axes is always a \textit{post-hoc} procedure and may have little relation to the overall question (for example, if the first few ordination axes represent elongation of the beak in birds, but the question is about wing disparity).
Additionally, in some cases, reducing the dimensionality of a dataset can render its interpretation more problematic.
For example, when the analysed data are non-Euclidean (e.g. some types of discrete morphological characters such as inapplicable characters), interpreting the resulting ordinated space can be challenging \citep{Gerber2019}.
\textcolor{blue}{COLORTAGIN[This can be sometimes problematic when comparing the position of groups in multidimensional space, as the distances might not be linear (although this can sometimes be corrected ]COLORTAGOUT \citealt{cailliez1983})}.
Furthermore, \textit{post-hoc} interpretations of the gradient of variation on the ordination axes may be biologically meaningless or simply impossible \citep{Gerber2019}.
Although some gradients are easy to detect or interpret (e.g.
the elongation and depth of mandibles in fishes on first and second PC axes, respectively; \citealt{Hill2018}), some are not (e.g. \citealt{Weisbecker2019}).
For example, with discrete morphological data, a gradient between the species that have many characters in state 1 and those that have more in state 0 has no biological meaning if these are binary alternate states.

@@

\begin{figure}[!htbp]
\centering
   \includegraphics[width=0.9\textwidth]{Figures/figure_space2.pdf}
\caption{
    \textcolor{blue}{COLORTAGIN[Illustration between the different morphospaces and visualisation of the same dataset (the classic ``iris'' dataset of]COLORTAGOUT} \citealt{edgar1935irises,fisher1936use}).
    \textcolor{blue}{COLORTAGIN[Morphospaces: different mathematical representations of a morphospace. A trait matrix can be an ordinated matrix (e.g. in }\citealt{tyler2011detecting}\textcolor{blue}{COLORTAGIN[ ) or transformed into a distance matrix (e.g. in }\citealt{Close2015}\textcolor{blue}{COLORTAGIN[, not represented here).}
    \textcolor{blue}{COLORTAGIN[Here we consider all these matrices as being \textit{morphospaces}, i.e. objects containing all the combinations of traits and observations (albeit transformed differently).
    Visualisation: different  ways to represent the morphospace in 2D.
    Visualisations can use either trait plots (directly from the trait matrix); or ordination axis plots (directly from the ordinated matrix).
    Note that in 2D representations, it is good practice to plot both axes on the same scale to avoid visually distorting the importance of one axis).]COLORTAGOUT}
}
\label{Fig:morphospace}
\end{figure}

\textcolor{blue}{COLORTAGIN[\subsection{Summarising disparity using disparity indices}]COLORTAGOUT \label{section:metrics}}

@@

When considering only one dimension, \textcolor{blue}{COLORTAGIN[disparity indices can be used to compare the spread of distributions (e.g. the range, quantiles or variance) or the differences in the central tendencies (i.e. mean, median or mode) of groups in the morphospace.]COLORTAGOUT}
Among these indices, some will have more attractive properties than others, such as sensitivity to outliers.
Range and mean are highly sensitive, whereas quantiles, variance and median are less so, making them more or less appropriate for different questions.
For example, if the goal is to characterise the extent of morphospace occupied by a group (e.g. does group A occupy as much space as group B?), indices related to the spread of the group in the morphospace are most appropriate (e.g.
volume; \citealt{Diaz2016}, distance from the centroid; \citealt{Hopkins2017, Finlay2015}, variance and range; \citealt{Brusatte2008}).
Furthermore, aspects other than variation (\textit{sensu} disparity) can be of interest: if we wish to describe the ``position'' of a group in a morphospace (e.g.
does group A occupy the same space as group B?), indices related to the distance between the elements within a group and a fixed point in the morphospace are most appropriate \citep{GuillermeMOMS}.
Finally, if we aim to characterise the density of morphospace occupation (e.g.
is group A more closely packed than group B?) indices related to the pairwise distances between elements will be most appropriate (e.g. nearest neighbour distance, pairwise distances, etc. \citealt{Close2015} - see section \ref{section:ordination} above).

@@

\textcolor{blue}{COLORTAGIN[\subsection{Testing for differences in disparity} ]COLORTAGOUT\label{section:testing}}

No matter which disparity indices have been calculated, the research question must be framed in an appropriate statistical context.
The multidimensional statistical toolkit for ecology and evolution has greatly expanded in recent years \citep{clavel2015mvmorph, Adams2018}, but some of these advances have yet to be implemented in disparity analyses.
Instead, hypothesis testing has been mostly confined to a small set of well-established methods.
One commonly used test is the non-parametric permutation analysis of variance (PERMANOVA) \citep{Anderson2001} that tests whether two groups share the same centroid and dispersion based on a distance matrix between observations.
\textcolor{blue}{COLORTAGIN[The last decade has also seen a series of developments based on this test (e.g. the linear regression for multidimensional data \citealt{collyer2015method} or the phylogenetic ANOVA \citealt{Adams2014}; but see \citealt{Adams2018,lloyd2016} for more).
It is worth noting that most of these test do not require the morphospace to be ordinated (see section \ref{section:ordination} above).
Regardless of the statistical test used, they should only be employed if they are tailored to the question at hand, rather than simply following common practices.]COLORTAGOUT}

It is also important to consider which data should be subjected to a statistical test.
For example, in morphological disparity analysis, especially for palaeobiological questions, data are often bootstrapped.
This has two advantages: (i) when the disparity index is unidimensional (e.g. the sum of variances), bootstrapping the data generates a distribution of the index that can be analysed using the vast statistical toolkit available for comparing distributions; (ii) when data are scarce, bootstrapping the data allows users to introduce variance, rendering the test less sensitive to outliers.
However, bootstrapped data are pseudoreplicates and thus non-independent \textcolor{blue}{COLORTAGIN[and can increase the false positive rate (Type I error)]COLORTAGOUT} \citep{strube1988bootstrap}.
\textcolor{blue}{COLORTAGIN[This, again, highlights the importance of tailoring the statistical test to the data and question at hand.]COLORTAGOUT}

@@

As with all comparative datasets, the data used in disparity analyses are not independent because close relatives will tend to have more similar morphologies than more distant relatives \citep{Harvey1998}.
Thus, for disparity analyses that consider groups with phylogenetic relationships (which is common), the non-independence between observations should be taken into account.
It has been noted, however, that some popular phylogenetic correction methods (like phylogenetic PCA) can be inappropriate, especially when using only the first \textit{d} axes of the ordination, and can lead to incorrect interpretations of the data (such as wrongly supporting ``early burst'' type patterns; \citealt{Uyeda2015}).
\textcolor{blue}{COLORTAGIN[Furthermore, any use of phylogenies in disparity analyses must also carefully consider the underlying model of trait evolution. Standard methods assume a model of Brownian motion, i.e. a ``random walk'' model where trait variance increases linearly through time with no trend in the direction of trait evolution. In many biological situations this model is not realistic, and different models of evolution should be considered \citealt{blomberg2020}. If an inappropriate model is used then methods such as phylogenetic PCA and ancestral state estimations (see below) may give misleading results, with implications for downstream results of disparity analyses.]COLORTAGOUT}

One other common way to take phylogeny into account in disparity analyses is using ancestral state estimations in \textcolor{blue}{COLORTAGIN[disparity through time]COLORTAGOUT} analyses to extract disparity estimates for non-sampled taxa and/or nodes of a phylogeny \citep{brusatte2011phylogenetic,Guillerme2018}.
Ancestral state estimation can be performed at two points in the disparity analysis pipeline: either (1) pre-transformation, i.e. the estimation is done before transformation of the data (e.g. ordination, or distance matrix construction) and is simply based on the original data; or (2) post-transformation, i.e. the estimation is done after transformation of the data by estimating the ancestral states using the transformed matrix (e.g. the ordination scores; \citealt{lloyd2018}).
